A Linear Solution of the Linearity Challenge

### 2025-12-28 paper/Examples.lagda.tex
tuning
remark in example and revised related

\begin{code}
!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
!!_ B = curry∗ cut B ≫ (If False Else True)

_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
A && B   =  curry∗ cut A ≫ $
            curry∗ cut B ≫ $
            If (curry∗ link ch (< ≫) ch) Else (Drop False)
A || B   = !! ((!! A) && (!! B))
\end{code}

The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
visible function application operator. We use it as a ``separator'' so that we
can flatten deeply nested expressions and save a bunch of parentheses.
%
We have chosen to define the disjunction \AgdaFunction{||} from the conjunction
\AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's laws.

If we wish to test our definitions, we can implement a simple evaluator using
the deadlock freedom property. We have not proved a termination result, but
since linear logic enjoys cut elimination we can safely annotate the evaluator
as terminating.

@@

The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and
creates an element of \AgdaDatatype{Fin} from the natural number $n$.
%
We now encode the server
\begin{code}
Server : Proc [ ServerT ]
Server = curry (curry∗ server ch (< ≫)) un-[] $
         curry∗ all ch (< ≫) λ X →
         curry∗ join ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
         close ch
\end{code}
and the client that sends true to it
\begin{code}
Client : Proc (dual ServerT ∷ 𝔹 ∷ [])
Client = curry∗ client ch (< ≫) $
         curry∗ (ex {_} {𝔹}) ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) True ≫ $
         curry∗ join ch (< ≫) $
         curry∗ wait ch (< ≫) $
         curry∗ link ch (< > •) ch
\end{code}

We define a main process that connects server and client
\begin{code}
Main : Proc [ 𝔹 ]
Main = curry∗ cut Client (< •) Server
\end{code}
and then ask Agda to normalize \AgdaFunction{Main}, which yields
\AgdaFunction{True} as expected.

>>>>

\begin{code}
!! : Proc [ 𝔹 ] → Proc [ 𝔹 ]
!! B = curry∗ cut B ≫ (If False Else True)

_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
A && B   =  curry∗ cut A ≫ $
            curry∗ cut B ≫ $
            If curry∗ link ch (< ≫) ch Else (Drop False)
A || B   =  !! (!! A && !! B)
\end{code}

The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
low-precedence, visible function application operator. We use it as a separator
to flatten deeply nested expressions and save a bunch of parentheses.
%
For the sake of illustration, we have chosen to define the disjunction
\AgdaFunction{||} from the conjunction \AgdaFunction{\&\&} and negation
\AgdaFunction{!!} using De Morgan's laws.

To test our definitions we implement a simple evaluator using the deadlock
freedom property. We have not proved a termination result, but since linear
logic enjoys cut elimination we can safely annotate the evaluator as
terminating.

@@

The notation $\AgdaFunction{\#}~n$ (defined in Agda's standard library) creates
an element of \AgdaDatatype{Fin} from the natural number $n$. Here it is used to
create the de Bruijn index of the type variable $X$.
%
We now encode the server
\begin{code}
Server : Proc [ ServerT ]
Server = curry (curry∗ server ch (< ≫)) un-[] $
         curry∗ all ch (< ≫) λ X →
         curry∗ join ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
         close ch
\end{code}
and the client that sends true to it
\begin{code}
Client : Proc (dual ServerT ∷ 𝔹 ∷ [])
Client = curry∗ client ch (< ≫) $
         curry∗ (ex {_} {𝔹}) ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) True ≫ $
         curry∗ join ch (< ≫) $
         curry∗ wait ch (< ≫) $
         curry∗ link ch (< > •) ch
\end{code}

To test our definitions we compose client and server in parallel
\begin{code}
Main : Proc [ 𝔹 ]
Main = curry∗ cut Client (< •) Server
\end{code}
and then ask Agda to normalize \AgdaFunction{Main}, which yields
\AgdaFunction{True} as expected.

\begin{remark}
    Writing processes in Agda would be easier if we could avoid using
    \AgdaFunction{curry∗}, namely if the constructors of the data type
    \AgdaDatatype{Proc} were naturally curried.
    %
    This is possible, but only if we are willing to sacrifice some clarity. To
    illustrate the problem with an example, below is the fully curried
    constructor \AgdaInductiveConstructor{fork} of a hypothetical data type
    \AgdaDatatype{Proc'}, obtained by expanding the definition of separating
    conjunction:
\begin{code}[hide]
data Proc' : Context → Set where
\end{code}
\begin{code}
  fork : ∀{A B Γ Δ Θ Θ₁ Θ₂} → Ch (A ⊗ B) Δ → Γ ≃ Δ + Θ →
         Proc' (A ∷ Θ₁) → Θ ≃ Θ₁ + Θ₂ → Proc' (B ∷ Θ₂) → Proc' (A ⊗ B ∷ Γ)
\end{code}

    This version of \AgdaInductiveConstructor{fork} is way less readable than
    the one we gave in \Cref{sec:process-agda} because of the (now visible)
    context splittings. We can recover some clarity and still obtain a curried
    version of \AgdaInductiveConstructor{fork} using the magic wand instead of
    Agda's native arrow:
\begin{code}[hide]
data Proc'' : Context → Set where
\end{code}
\begin{code}
  fork : ∀{A B} →
         ∀[ Ch (A ⊗ B) ⇒ (A ∷_) ⊢ Proc'' ─∗ (B ∷_) ⊢ Proc'' ─∗ Proc'' ]
\end{code}

    However, now in each constructor we have to make a distinction between the
    first arrow \AgdaFunction{⇒} and the following magic wands \AgdaFunction{─∗}
    to account for the appropriate amount of context splittings. We found this
    formulation of the constructors harder to explain and motivate in
    \Cref{sec:process-agda}.
    %
    In the end, none of the alternative definitions of the data type
    \AgdaDatatype{Proc} was fully satisfactory and we preferred the cleanest and
    most readable version at the expense of additional clutter in this section.
    %
    \eoe
\end{remark}

### 2025-12-28 paper/abstract.tex
remark in example and revised related

\abstract{%
    The formalisation of session calculi is considered challenging because of
    the management of channels, which are linear resources that cannot be
    discarded or duplicated and whose type changes over time, as input/output
    operations are performed on them.
    %
    Context splitting, the ``naive'' channel management technique directly
    related to the way session type theories are usually written with pen and
    paper, is often considered a hindrance and a notable source of complexity,
    to the point that several alternative approaches have been recently
    proposed.
    %
    In this paper we describe the Agda formalisation of a process calculus based
    on classical linear logic that supports the modeling of binary sessions
    through their encoding with explicit continuation channels. The
    formalisation turns out to be remarkably compact despite the adoption of
    context splitting. We argue that the logical nature of the calculus and the
    use of explicit continuations are contributing factors to the simplicity of
    its formalisation.
    %
}

>>>>

\abstract{%
    The formalisation of session calculi is considered challenging because of
    the management of channels, which are linear resources that cannot be
    discarded or duplicated and whose type changes over time, as input/output
    operations are performed on them.
    %
    Context splitting, the channel management technique directly related to the
    way session type theories are usually written with pen and paper, is often
    considered a hindrance and a notable source of complexity, to the point that
    several alternative approaches have been recently proposed.
    %
    In this paper we describe the Agda formalisation of a process calculus based
    on classical linear logic that supports the modeling of binary sessions
    through their encoding with explicit continuation channels. The
    formalisation turns out to be remarkably compact despite the adoption of
    context splitting. We argue that the logical nature of the calculus and the
    use of explicit continuations are contributing factors to the simplicity of
    its formalisation.
    %
}

### 2025-12-28 paper/related.tex
remark in example and revised related

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Looking on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the features of \CP.

The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are the
ones that adopt a locally nameless representation of channels. In these
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

>>>>

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors.
%
% Also, most formalisations predate the publication of the linearity
% challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
% solving that particular problem. However, some of
% them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
% are explicitly motivated by the complexity arising from the linear usage of
% resources/channels and from context splitting in particular.
%
Looking at the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the features of \CP.

The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are
also the ones that adopt a locally nameless representation of channels. In these
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This entails some duplication of effort as well
as some transformation machinery between the two representations.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.
